Identifying Shifts in the Share Distribution of Transboundary Fish Stocks to Lead Time for Policy Response Under Climate Change

### 2020-11-24 Scripts/Manuscript.Rmd
misc

This analysis includes 633 exploited marine transboundary species previously identified to account for 80% of the catch taken from the world’s EEZs between 2005 and 2014 [@PalaciosAbrantes:2020tv]. Each time a species was shared by a pair of neighbouring EEZs it was considered a transboundary stock [@Teh:2015gd; @PalaciosAbrantes:2020tv] for a total of 9,132 transboundary stocks. We defined the boundaries of the world’s EEZs using the Sea Around Us spatial division (updated 1 July 2015, available from http://www.seaaroundus.org), noting that it subdivides the EEZs of 198 coastal states into 280 regions (e.g., Mexico’s EEZ is divided into Mexico Pacific and Mexico Atlantic; Figure \@ref(fig:FigureS1)) including island territories. We determined the intersections between polygons using the R-statistical software package sf [@PackagesfSimpleF:2018vp]. Each EEZ was categorized by the geopolitical regions according to the United Nations and by natural biome [@Reygondeau:2019uh]. The habitat association of each species was determined based on information available from FishBase (http://www.fishbase.org). For each species and EEZ, we used the Sea Around Us data to estimate catch and fishing revenue from fishing activities within global EEZs. We report both average catch and revenue for the last decade (2005-2014) [@Zeller2016; @Tai2017; @Sumaila:2015uc]. The monthly average US Consumer Price Index (CPI) according to the U.S. Bureau of Labor Statistic (https://www.bls.gov/cpi/) was used to standardize all monetary values to 2019 real USD.

@@

We developed an index to evaluate range shifts in transboundary stocks under climate change (TI). This index is based on the shifts in the distribution centroid of a transboundary stock relative to the centroid of the neighboring EEZs that share this stock (Figure \@ref(fig:FigureS2)). The centroid of a transboundary stock was determined by the average latitude ($lat_{ts}$) and longitude ($lon_{ts}$) across the grid cells with the highest abundance within the neighboring EEZs sharing the stock e.g., sockeye salmon (*Oncorhynchus nerka*) that is transboundary between the United States and Canada. Therefore,

@@

where $lat_{per}$ and $lon_{per}$ are the latitudes and longitudes of the grid cells holding the $per^{th}$ percentile of the projected transboundary stock abundance. To focus on areas where transboundary stocks are more abundant and fishing activities are more likely to take place, we included grid cells where the projected stock abundance within the neighbouring EEZs sharing the stock was above the top $10^{th}$ percentile (e.g., per = 10%). We ran a sensitivity analysis (e.g. per = $20^{th}$, $50^{th}$, $90^{th}$ percentile) and examined the effects of different thresholds on the calculated index value (Figure \@ref(fig:FigureS2)). The centroid of a EEZ was estimated using the st package in R (version 3.2.1) (Figure \@ref(fig:FigureS1)). For each ensemble member, neighboring EEZs and transboundary stock, we computed the distance between centroids assuming the earth is a perfect sphere and ignoring geographic barriers using the geosphere package in R;

@@

The average ToE of the transboundary index (TI) across the emerging stocks in all EEZs analyzed is projected to be 2029 $\pm$ 26 years (Figure \@ref(fig:Figure1). Fish stocks with TI that first emerged from their historical variability in each EEZ was projected to be as early as 2006. In total, the projected transboundary index of 5,745 stocks (63% of the studied stocks) will emerge from their historical variability between 2006 and 2100. About 55% (n = 3,154) of those stocks are projected to have their shifts in distributions emerged between 2006 and 2020. Also, 96% of the world’s EEZs saw at least one stock having their distribution shifts between 2006 and 2020 (e.g. *Engraulis ringens* shared by Chile and Peru emerged in 2010). Shifts in stocks’ distributions will continue to exceed historical limits steadily towards the end of the $21^{st}$ century and are not expected to stop there (Figure \@ref(fig:Figure1)). The number of EEZs with emerging stocks is projected to increase steeply from 87% to 96% during the period of 2006 to 2020. After the mid-2020s, the rate of increase in the number EEZs slow down to an average of 0.74% per year. By 2081, all of the EEZs analyzed here will have at least one emerging stock (Figure \@ref(fig:Figure1)).

```{r Figure1, eval = T,  echo = F, fig.width = 9, fig.height=4, fig.pos= "H", fig.cap =  "Time of emergence by Exclusive Economic Zone and transboundary stock. A) Showing results in absolute numbers. B) Showing the result as a percentage of the total stocks and EEZ."}

@@

The median ToE of the transboundary index of stocks varied significantly according to the geographic region of the neighboring EEZs (Kruskal-Wallis, X2 = 287.23, DF = 93, p < 0.001). Around one fifth (19%) of the EEZs have an average time of emergence across stocks prior to 2020 (Figure 2). Notable exceptions where the ToE is relatively late include Honduras (ToE = 2071), New Zealand (ToE = 2055), Ireland (ToE = 2067) and Pitcairn (ToE = 2070). The EEZs in Eastern Asia, Latin America, the Caribbean and Polynesia regions differed significantly (p < 0.05) from (almost) all other regions (see Table S1 for test statistics). (Figure \@ref(tab:TableS1).

Tropical latitudes also hold the majority of countries with the largest number of shifts in transboundary stocks due to climate change (Figure \@ref(fig:Figure2). Only in very few cases like Brazil and Alaska’s and Russia’s polar EEZs, no stock will emerge between 2006 and 2100. From the top 10 more vulnerable countries, seven are located in tropical waters (Tokelau 100, the	Marshall Islands 92, Peru 88, Kiribati 80, Tuvalu 80, Papua New Guinea 78, and the Bahamas 74) while only three are outside the tropics (Estonia 86, Finland 82 and Sweden 78). We set an arbitrary threshold of  1 s.d. (67% confidence) to determine the ToE (see Methods). We tested the sensitivity of our results using a larger threshold (i.e.,  2 s.d for a 95% confidence). While the selected 1 s.d. threshold is less conservative than a 2 s.d., the distribution of results do not dramatically change between thresholds, with a difference in ToE of about 4 years (Figure \@ref(fig:FigureS2)).

```{r Figure2, eval = T,  echo = F, fig.width  = 13,fig.pos= "H", fig.height =11, fig.cap =  "Average time of emergence of transboundary stocks. A) Land polygon shows the vulnerability index of individual country’s fisheries management to shifts in transboundary stocks. The warmer the colour the greater the vulnerability. Exclusive Economic Zone polygon represents the average time of emergence across all stocks, with the cooler the colour the later the average time of emergence. The Density plot on the right shows the latitudinal number of emerging stocks across EEZ centroids (10 degrees latitude running mean). Aqua color = no emerging stocks. B) Time of emergence according to the United Nations sub regions. Whiskers represent 1.5* interquartile range. Box represents interquartile range as distance between first and third quartiles. Line represents median, and black points represent points outside of 1.5*IQR."}

@@

<!-- ![Figure 2. Average time of emergence of transboundary specie. Land polygon shows the vulnerability index. Exclusive Economic Zone polygon represents the average time of emergence across all stocks. Density plot on the right shows the latitudinal number of emerging stocks across EEZ centroids (10^o running mean).](../Figures/Fig3.png) -->

Our analysis suggests that most (n = 47) countries responsible for 75% of the yearly revenue generated from transboundary species between 2005 and 2010 will see their distribution range shifts beyond historical variabilities before 2040 (Figure \@ref(fig:Figure3)). Particularly, nine of the top ten countries earning the most revenue from transboundary species, share a border. In some cases, like Indonesia, Thailand and the United States, stocks’ range shifts swill emerge by around 2030. The 47 countries that make up the bulk of the global fisheries revenues also have more stocks with range shifts emerging from historical variabilities.

```{r Figure3, eval = T,  echo = F, fig.width=7, fig.height=8,fig.pos= "H", fig.cap =  "Average time of emergence of transboundary species per fishing entity, aggregated by colour according to region. Fishing revenue on a logarithmic scale. Showing country names for the top 90th percentile of annual fishing revenue."}

@@

<!-- ![Figure 3. Average time of emergence of transboundary species per fishing entity. Fishing revenue in logarithmic scale. Showing names for the top 75th percentile. Countries marked with an asterisk have at least one neighboring border in the group.](../Figures/Fig4.png) -->

We compared species time of emergence within EEZs by species’ habitat association (Figure \@ref(fig:Figure4)). Each species is assigned to be associated to one of the xx habitat types based on information available from FishBase (for fish, ref) or SeaLifeBase (for invertebrates, ref). We found significant differences in the time of emergence of stocks; range shifts by their habitat association (Kruskal-Wallis, $X^2$ = 286.48, DF = 93, p < 0.001). Specifically, time of emergence of range shifts of species that are associated with pelagic-oceanic and reef habitats were significantly different than other habitat preferences (See Table \@ref(tab:TableS2)). On average, species associated with pelagic oceanic habitats e,g, tunas (*Thunnus sp.*), bathy-pelagic habitats e.g., blue whiting (*Micromesistius poutassou*) and corals (e.g., Bigeye snapper *Lutjanus lutjanus*) are projected to have an earliest time of emergence (Figure \@ref(fig:Figure4)). Species associated with bathydemersal habitats such as Alaska plaice (*Pleuronectes quadrituberculatus*) generally have an average time of emergence of range shifts that is later than 2025 (Figure \@ref(fig:Figure4)).

```{r Figure4, eval = T,  echo = F, fig.width=14, fig.height=14,fig.pos= "H", fig.cap =  "Regional time of emergence per habitat association. Regions according to the United Nations and habitat association to FishBase. Whiskers represent 1.5* interquartile range. Box represents interquartile range as distance between first and third quartiles. Line represents median, and black points represent outliers outside of 1.5*IQR."}

@@

<!-- ![Figure 4. Regional time of emergence per habitat association. Regions according to the United Nations and habitat association to FishBase. Whiskers represent 1.5* interquartile range. Box represents interquartile range as distance between first and third quartiles. Line represents median, and black points represent outliers (outside of 1.5*IQR.](../Figures/Fig5.png) -->

<!-- We hypothesized that species environmental preferences (e.g. presence in certain biomes and habitats), climate change effects on water temperature (e.g. ToE of SST), human drivers (e.g. catch rate), and latitudinal gradient would have significant influence in the ToE of transboundary species (Table \ref(tab:TableS3)). Results from the multiple linear regression analysis show a significant relation between ToE of transboundary species and latitudinal gradient (*p* > 0.001), tropical (*p* = 0.013) and temperate (*p* = 0.002) biomes, EEZs in the Americas (*p* > 0.001), Asia (*p* > 0.001) and Oceania (*p* = 0.002), and some habitats (*p* > 0.001) (Table \ref(tab:Table1)). -->

@@

We estimated the changes in the SSR of transboundary stocks by the early (2020-2040) and the mid $21^{st}$ century (2040-2060), relative to the recent past (1951-2005) and highlighting changes below a neighbouring country’s TP. Our results suggest that 87% (n = 238) of the world EEZs will see a change in SSR below the country’s TP by 2030 (Figure \@ref(fig:Figure5)A). Globally, average changes in SSR of transboundary stocks in most EEZs are projected to be lower than 25% by the early $21^{st}$ century. The changes in SSR is expected to increase slightly towards the mid $21^{st}$ century (Figures \@ref(fig:FigureS4) and \@ref(fig:FigureS5)). However, changes in SST for specific stocks and in some EEZs are projected to be exceptionally high. For example, Guatemala is sharing *Panulirus gracilis* stock with Mexico, and Guatemala’s share is projected to increase by >50% from 12% to 26% by the early $21^{st}$ century. By the early $21^{st}$ century, the percentage of transboundary stocks per EEZ that will see a change in SSR above the country’s TP historical distribution  (e.g., $\Delta{SSR} > \mu \pm \sigma$) will be relatively large (global mean of 69% of stocks in an EEZ).

```{r Figure5, eval = T,  echo = F, fig.width=12, fig.height=10,fig.pos= "H", fig.cap =  "Changes in Specie-Share Ratio by 2030 (2021-2040) relative to 1951-2005. Lines represent the average gain of transboundary stock share ratio with arrows going from EEZ with decreasing stock share (point) to those that are gaining shares (arrowhead). Land polygons represent the percentage of initial stocks that will see a shift. Panel B zooms in to specific areas shown within the grey dashed line boundaries in A."}

@@

In this paper we endeavor to look at the timing and intensity of changes in the sharing of transboundary species. Our findings highlight the early emergence of transboundary distribution shifts from historical variations for most of the shared stocks in the world. Such early emergences of range shifts indicate that the immediate challenges climate change is posing on international fisheries management (Figure \@ref(fig:Figure1)). Moreover, we identify regional patterns of high climate risk to important fish stocks and associated fisheries at the present-day that will require fast responses from transboundary fisheries management \@ref(fig:Figure2) and \@ref(fig:Figure3)). Finally, we show that in many instances around the world, the coming years will see that the shares of many transboundary stocks by countries will drop below level that conduce cooperation between countries in managing the shared stocks (Figure \@ref(fig:Figure5)).

@@

Our results suggest that in most EEZs, more than half of their transboundary stocks emerged in the early 2000s (Figure \@ref(fig:Figure1)). Our projections in the emergence of distribution shifts of transboundary stocks corroborate with previous studies that have detected and attributed changes in marine catch composition since the 1970s [@Cheung:2013gk; @Poloczanska:2016kk]. For example, in the early 2000s, Humboldt squid (*Dosidicus gigas*) substantially expanded its geographic range poleward, reaching the coast of Washington state (USA). A new fishery targeting Humboldt squid was developed quickly following its range expansion [@Pinsky:2014by; @Zeidberg:2007gs]. In the North East Atlantic, the range expansion of Atlantic mackerel (*Scomber scombrus*) into Icelandic waters in 2010 was followed by the increase in fishery quota **xxx** without consultation with **xxxx.** Such changes in quota resulted in disputes Iceland and Faroes Islands with multiple fishing nations co-managing the Atlantic mackerel stocks [@Spijkers:2017ij].

@@

Our results underscore regional ‘hotspots’ of climate risk for transboundary fisheries management (Figure \@ref(fig:Figure2)). The premise of our vulnerability index is that, from a management perspective, countries whose shifting transboundary stocks provide larger societal benefits and have an early ToE will be more vulnerable to shifts in SSR than their counterpart. Thus, climate risk hotspots for transboundary management happen where EEZs will see an early ToE and where transboundary stocks are important in terms of fisheries catches and revenues (Figure \@ref(fig:Figure2)). For example, Polynesia and Micronesia are the only regions where the average ToE across all stocks and EEZs already happened; 2019 or is about too 2021, respectively (Figure \@ref(fig:Figure5)). Countries in these regions are highly dependent on pelagic-oceanic (e.g.,tunas and bill fishes; [@Johnson:2017ev]) and reef-associated transboundary stocks [@Cabral:2018hc] for both food security and profits [@CisnerosMontemayor:2016gq; @Hanich:2018dh]. For example, in the Marshall Islands 16 transboundary stocks responsible for 92% and 94% of the transboundary fisheries revenue and catch, respectively, have, on average, emerged since 2018 resulting in high vulnerability index (Figure \@ref(fig:Figure5)). In this case, policy has to work against the clock and under potentially high societal pressure to address an issue that could be impacting important fisheries [@Burden:2019ij]. Currently, fisheries targeting shared stocks like tunas in the South Pacific Islands region operate under the Vessel Day Scheme - which takes into account shifts in stocks and catches making it a good example where a management system has been set up that is somewhat anticipatory [@Aqorau:2018bh]. However, a low vulnerability index does not necessarily translate to no-risk of international conflict or threat to stocks sustainability. While it gives countries more time to prepare for substantial changes, stocks with relatively small overall participation on catches or revenue can destabilize cooperative shared stock management agreement as shown by the Atlantic mackerel example [@Spijkers:2017ij, Chapter 1].

@@

Changes in SSR will be seen across the world, suggesting that climate change impacts on transboundary stock management is a global problem (Figure \@ref(fig:Figure5)). Poleward shifts in SSR are expected in regions of the world with relatively large EEZs and few boundaries, like the North East Pacific. However, despite observations and modelling work indicating that, generally, climate change is shifting stocks distributions poleward [@Cheung:2010dt; @Poloczanska:2016kk; @IPCC:2019tn; @Fogarty:2017eo], the Eastern Tropical Pacific will see EEZs from Mexico to Ecuador losing SSR to their southern neighbors. Such range shifts towards the equator reflect movements to cooler habitats characteristic of equatorial upwelling systems. They are consistent with regional modeling studies suggesting that by mid $21^{st}$ century, marine stocks will move between 50 and 100 km southward of Mexico’s EEZ towards Peru under a high emission climate change scenario [@Clarke:2020ep].  Moreover, regions characterized by multiple political boundaries, like the Caribbean and the North East Atlantic, will see a complex network of changes in SSR with no clear pattern (Figure \@ref(fig:Figure5)B). These regions are known for being highly connected by both fish adult migration [@Levin:2018in] and larvae dispersal [@Ramesh:2019va] supporting the implementation of joint management plans [@Baudron:2020dj] and the call for new implemented shared policies [@Levin:2018in]. However, game theory suggests that the greater the number of negotiating parties , the harder it is for parties to reach an agreement [@Sumaila:2013hv], thus, making it particularly challenging for countries in these regions to coordinate the management of shifting shared stocks.

@@

## Supplements

```{r FigureS1, eval = F,  echo = F, fig.width=9, fig.height=4, fig.cap =  "World Exclusive Economic Zones used in this study as devided by the Sea Around Us and their centroids (points)."}

@@

```{r FigureS2, eval = F,  echo = F, fig.width=9, fig.height=4, fig.cap = "Graphical representation of the transboundary index (TI) to determine the time of emergence of transboundary stocks. The index is based on the distance between the distributional centroid of the transboundary stock and the geographic centroid of the neighbouring Exclusive Economic Zones sharing the stock (Top panel). Time of emergence is defined as the first year when the TI overpasses historical values (Bottom panel)."}

@@

```{r FigureS3, eval = F,  echo = F, fig.width=4, fig.height=4, fig.cap =  "Average Time  of Emergence and standar dviation for a 64% confidence treshold  (purple) and a 98% confidence teshold (yellow)"}

>>>>

This analysis includes 633 exploited marine transboundary species previously identified to account for 80% of the catch taken from the world’s EEZs between 2005 and 2014 [@PalaciosAbrantes:2020tv]. Each time a species was shared by a pair of neighbouring EEZs it was considered a transboundary stock [@Teh:2015gd; @PalaciosAbrantes:2020tv] for a total of 9,132 transboundary stocks. We defined the boundaries of the world’s EEZs using the Sea Around Us spatial division (updated 1 July 2015, available from http://www.seaaroundus.org), noting that it subdivides the EEZs of 198 coastal states into 280 regions (e.g., Mexico’s EEZ is divided into Mexico Pacific and Mexico Atlantic; Figure \@ref(fig:toeFigureS1)) including island territories. We determined the intersections between polygons using the R-statistical software package sf [@PackagesfSimpleF:2018vp]. Each EEZ was categorized by the geopolitical regions according to the United Nations and by natural biome [@Reygondeau:2019uh]. The habitat association of each species was determined based on information available from FishBase (http://www.fishbase.org). For each species and EEZ, we used the Sea Around Us data to estimate catch and fishing revenue from fishing activities within global EEZs. We report both average catch and revenue for the last decade (2005-2014) [@Zeller2016; @Tai2017; @Sumaila:2015uc]. The monthly average US Consumer Price Index (CPI) according to the U.S. Bureau of Labor Statistic (https://www.bls.gov/cpi/) was used to standardize all monetary values to 2019 real USD.

@@

We developed an index to evaluate range shifts in transboundary stocks under climate change (TI). This index is based on the shifts in the distribution centroid of a transboundary stock relative to the centroid of the neighboring EEZs that share this stock (Figure \@ref(fig:toeFigureS2)). The centroid of a transboundary stock was determined by the average latitude ($lat_{ts}$) and longitude ($lon_{ts}$) across the grid cells with the highest abundance within the neighboring EEZs sharing the stock e.g., sockeye salmon (*Oncorhynchus nerka*) that is transboundary between the United States and Canada. Therefore,

@@

where $lat_{per}$ and $lon_{per}$ are the latitudes and longitudes of the grid cells holding the $per^{th}$ percentile of the projected transboundary stock abundance. To focus on areas where transboundary stocks are more abundant and fishing activities are more likely to take place, we included grid cells where the projected stock abundance within the neighbouring EEZs sharing the stock was above the top $10^{th}$ percentile (e.g., per = 10%). We ran a sensitivity analysis (e.g. per = $20^{th}$, $50^{th}$, $90^{th}$ percentile) and examined the effects of different thresholds on the calculated index value (Figure \@ref(fig:toeFigureS2)). The centroid of a EEZ was estimated using the st package in R (version 3.2.1) (Figure \@ref(fig:toeFigureS1)). For each ensemble member, neighboring EEZs and transboundary stock, we computed the distance between centroids assuming the earth is a perfect sphere and ignoring geographic barriers using the geosphere package in R;

@@

The average ToE of the transboundary index (TI) across the emerging stocks in all EEZs analyzed is projected to be 2029 $\pm$ 26 years (Figure \@ref(fig:toeFigure1). Fish stocks with TI that first emerged from their historical variability in each EEZ was projected to be as early as 2006. In total, the projected transboundary index of 5,745 stocks (63% of the studied stocks) will emerge from their historical variability between 2006 and 2100. About 55% (n = 3,154) of those stocks are projected to have their shifts in distributions emerged between 2006 and 2020. Also, 96% of the world’s EEZs saw at least one stock having their distribution shifts between 2006 and 2020 (e.g. *Engraulis ringens* shared by Chile and Peru emerged in 2010). Shifts in stocks’ distributions will continue to exceed historical limits steadily towards the end of the $21^{st}$ century and are not expected to stop there (Figure \@ref(fig:toeFigure1)). The number of EEZs with emerging stocks is projected to increase steeply from 87% to 96% during the period of 2006 to 2020. After the mid-2020s, the rate of increase in the number EEZs slow down to an average of 0.74% per year. By 2081, all of the EEZs analyzed here will have at least one emerging stock (Figure \@ref(fig:toeFigure1)).

```{r toeFigure1, eval = T,  echo = F, fig.width = 9, fig.height=4, fig.pos= "H", fig.cap =  "Time of emergence by Exclusive Economic Zone and transboundary stock. A) Showing results in absolute numbers. B) Showing the result as a percentage of the total stocks and EEZ."}

@@

The median ToE of the transboundary index of stocks varied significantly according to the geographic region of the neighboring EEZs (Kruskal-Wallis, X2 = 287.23, DF = 93, p < 0.001). Around one fifth (19%) of the EEZs have an average time of emergence across stocks prior to 2020 (Figure \@ref(fig:toeFigure2). Notable exceptions where the ToE is relatively late include Honduras (ToE = 2071), New Zealand (ToE = 2055), Ireland (ToE = 2067) and Pitcairn (ToE = 2070). The EEZs in Eastern Asia, Latin America, the Caribbean and Polynesia regions differed significantly (p < 0.05) from (almost) all other regions (see Table S1 for test statistics). (Figure \@ref(tab:toeTableS1).

Tropical latitudes also hold the majority of countries with the largest number of shifts in transboundary stocks due to climate change (Figure \@ref(fig:toeFigure2). Only in very few cases like Brazil and Alaska’s and Russia’s polar EEZs, no stock will emerge between 2006 and 2100. From the top 10 more vulnerable countries, seven are located in tropical waters (Tokelau 100, the	Marshall Islands 92, Peru 88, Kiribati 80, Tuvalu 80, Papua New Guinea 78, and the Bahamas 74) while only three are outside the tropics (Estonia 86, Finland 82 and Sweden 78). We set an arbitrary threshold of  1 s.d. (67% confidence) to determine the ToE (see Methods). We tested the sensitivity of our results using a larger threshold (i.e.,  2 s.d for a 95% confidence). While the selected 1 s.d. threshold is less conservative than a 2 s.d., the distribution of results do not dramatically change between thresholds, with a difference in ToE of about 4 years (toeFigure \@ref(fig:toeFigureS2)).

```{r toeFigure2, eval = T,  echo = F, fig.width  = 13,fig.pos= "H", fig.height =11, fig.cap =  "Average time of emergence of transboundary stocks. A) Land polygon shows the vulnerability index of individual country’s fisheries management to shifts in transboundary stocks. The warmer the colour the greater the vulnerability. Exclusive Economic Zone polygon represents the average time of emergence across all stocks, with the cooler the colour the later the average time of emergence. The Density plot on the right shows the latitudinal number of emerging stocks across EEZ centroids (10 degrees latitude running mean). Aqua color = no emerging stocks. B) Time of emergence according to the United Nations sub regions. Whiskers represent 1.5* interquartile range. Box represents interquartile range as distance between first and third quartiles. Line represents median, and black points represent points outside of 1.5*IQR."}

@@

<!-- ![toeFigure 2. Average time of emergence of transboundary specie. Land polygon shows the vulnerability index. Exclusive Economic Zone polygon represents the average time of emergence across all stocks. Density plot on the right shows the latitudinal number of emerging stocks across EEZ centroids (10^o running mean).](../Figures/Fig3.png) -->

Our analysis suggests that most (n = 47) countries responsible for 75% of the yearly revenue generated from transboundary species between 2005 and 2010 will see their distribution range shifts beyond historical variabilities before 2040 (toeFigure \@ref(fig:toeFigure3)). Particularly, nine of the top ten countries earning the most revenue from transboundary species, share a border. In some cases, like Indonesia, Thailand and the United States, stocks’ range shifts swill emerge by around 2030. The 47 countries that make up the bulk of the global fisheries revenues also have more stocks with range shifts emerging from historical variabilities.

```{r toeFigure3, eval = T,  echo = F, fig.width=7, fig.height=8,fig.pos= "H", fig.cap =  "Average time of emergence of transboundary species per fishing entity, aggregated by colour according to region. Fishing revenue on a logarithmic scale. Showing country names for the top 90th percentile of annual fishing revenue."}

@@

<!-- ![toeFigure 3. Average time of emergence of transboundary species per fishing entity. Fishing revenue in logarithmic scale. Showing names for the top 75th percentile. Countries marked with an asterisk have at least one neighboring border in the group.](../Figures/Fig4.png) -->

We compared species time of emergence within EEZs by species’ habitat association (toeFigure \@ref(fig:toeFigure4)). Each species is assigned to be associated to one of the xx habitat types based on information available from FishBase (for fish, ref) or SeaLifeBase (for invertebrates, ref). We found significant differences in the time of emergence of stocks; range shifts by their habitat association (Kruskal-Wallis, $X^2$ = 286.48, DF = 93, p < 0.001). Specifically, time of emergence of range shifts of species that are associated with pelagic-oceanic and reef habitats were significantly different than other habitat preferences (See Table \@ref(tab:toeTableS2)). On average, species associated with pelagic oceanic habitats e,g, tunas (*Thunnus sp.*), bathy-pelagic habitats e.g., blue whiting (*Micromesistius poutassou*) and corals (e.g., Bigeye snapper *Lutjanus lutjanus*) are projected to have an earliest time of emergence (toeFigure \@ref(fig:toeFigure4)). Species associated with bathydemersal habitats such as Alaska plaice (*Pleuronectes quadrituberculatus*) generally have an average time of emergence of range shifts that is later than 2025 (toeFigure \@ref(fig:toeFigure4)).

```{r toeFigure4, eval = T,  echo = F, fig.width=12, fig.height=7,fig.pos= "H", fig.cap =  "Regional time of emergence per habitat association. Regions according to the United Nations and habitat association to FishBase. Whiskers represent 1.5* interquartile range. Box represents interquartile range as distance between first and third quartiles. Line represents median, and black points represent outliers outside of 1.5*IQR."}

@@

<!-- ![toeFigure 4. Regional time of emergence per habitat association. Regions according to the United Nations and habitat association to FishBase. Whiskers represent 1.5* interquartile range. Box represents interquartile range as distance between first and third quartiles. Line represents median, and black points represent outliers (outside of 1.5*IQR.](../Figures/Fig5.png) -->

<!-- We hypothesized that species environmental preferences (e.g. presence in certain biomes and habitats), climate change effects on water temperature (e.g. ToE of SST), human drivers (e.g. catch rate), and latitudinal gradient would have significant influence in the ToE of transboundary species (Table \ref(tab:toeTableS3)). Results from the multiple linear regression analysis show a significant relation between ToE of transboundary species and latitudinal gradient (*p* > 0.001), tropical (*p* = 0.013) and temperate (*p* = 0.002) biomes, EEZs in the Americas (*p* > 0.001), Asia (*p* > 0.001) and Oceania (*p* = 0.002), and some habitats (*p* > 0.001) (Table \ref(tab:toeTable1)). -->

@@

We estimated the changes in the SSR of transboundary stocks by the early (2020-2040) and the mid $21^{st}$ century (2040-2060), relative to the recent past (1951-2005) and highlighting changes below a neighbouring country’s TP. Our results suggest that 87% (n = 238) of the world EEZs will see a change in SSR below the country’s TP by 2030 (toeFigure \@ref(fig:toeFigure5)A). Globally, average changes in SSR of transboundary stocks in most EEZs are projected to be lower than 25% by the early $21^{st}$ century. The changes in SSR is expected to increase slightly towards the mid $21^{st}$ century (toeFigures \@ref(fig:toeFigureS4) and \@ref(fig:toeFigureS5)). However, changes in SST for specific stocks and in some EEZs are projected to be exceptionally high. For example, Guatemala is sharing *Panulirus gracilis* stock with Mexico, and Guatemala’s share is projected to increase by >50% from 12% to 26% by the early $21^{st}$ century. By the early $21^{st}$ century, the percentage of transboundary stocks per EEZ that will see a change in SSR above the country’s TP historical distribution  (e.g., $\Delta{SSR} > \mu \pm \sigma$) will be relatively large (global mean of 69% of stocks in an EEZ).

```{r toeFigure5, eval = T,  echo = F, fig.width=12, fig.height=10,fig.pos= "H", fig.cap =  "Changes in Specie-Share Ratio by 2030 (2021-2040) relative to 1951-2005. Lines represent the average gain of transboundary stock share ratio with arrows going from EEZ with decreasing stock share (point) to those that are gaining shares (arrowhead). Land polygons represent the percentage of initial stocks that will see a shift. Panel B zooms in to specific areas shown within the grey dashed line boundaries in A."}

@@

In this paper we endeavor to look at the timing and intensity of changes in the sharing of transboundary species. Our findings highlight the early emergence of transboundary distribution shifts from historical variations for most of the shared stocks in the world. Such early emergences of range shifts indicate that the immediate challenges climate change is posing on international fisheries management (Figure \@ref(fig:toeFigure1)). Moreover, we identify regional patterns of high climate risk to important fish stocks and associated fisheries at the present-day that will require fast responses from transboundary fisheries management \@ref(fig:toeFigure2) and \@ref(fig:toeFigure3)). Finally, we show that in many instances around the world, the coming years will see that the shares of many transboundary stocks by countries will drop below level that conduce cooperation between countries in managing the shared stocks (Figure \@ref(fig:toeFigure5)).

@@

Our results suggest that in most EEZs, more than half of their transboundary stocks emerged in the early 2000s (Figure \@ref(fig:toeFigure1)). Our projections in the emergence of distribution shifts of transboundary stocks corroborate with previous studies that have detected and attributed changes in marine catch composition since the 1970s [@Cheung:2013gk; @Poloczanska:2016kk]. For example, in the early 2000s, Humboldt squid (*Dosidicus gigas*) substantially expanded its geographic range poleward, reaching the coast of Washington state (USA). A new fishery targeting Humboldt squid was developed quickly following its range expansion [@Pinsky:2014by; @Zeidberg:2007gs]. In the North East Atlantic, the range expansion of Atlantic mackerel (*Scomber scombrus*) into Icelandic waters in 2010 was followed by the increase in fishery quota **xxx** without consultation with **xxxx.** Such changes in quota resulted in disputes Iceland and Faroes Islands with multiple fishing nations co-managing the Atlantic mackerel stocks [@Spijkers:2017ij].

@@

Our results underscore regional ‘hotspots’ of climate risk for transboundary fisheries management (Figure \@ref(fig:toeFigure2)). The premise of our vulnerability index is that, from a management perspective, countries whose shifting transboundary stocks provide larger societal benefits and have an early ToE will be more vulnerable to shifts in SSR than their counterpart. Thus, climate risk hotspots for transboundary management happen where EEZs will see an early ToE and where transboundary stocks are important in terms of fisheries catches and revenues (Figure \@ref(fig:toeFigure2)). For example, Polynesia and Micronesia are the only regions where the average ToE across all stocks and EEZs already happened; 2019 or is about too 2021, respectively (Figure \@ref(fig:toeFigure5)). Countries in these regions are highly dependent on pelagic-oceanic (e.g.,tunas and bill fishes; [@Johnson:2017ev]) and reef-associated transboundary stocks [@Cabral:2018hc] for both food security and profits [@CisnerosMontemayor:2016gq; @Hanich:2018dh]. For example, in the Marshall Islands 16 transboundary stocks responsible for 92% and 94% of the transboundary fisheries revenue and catch, respectively, have, on average, emerged since 2018 resulting in high vulnerability index (Figure \@ref(fig:toeFigure5)). In this case, policy has to work against the clock and under potentially high societal pressure to address an issue that could be impacting important fisheries [@Burden:2019ij]. Currently, fisheries targeting shared stocks like tunas in the South Pacific Islands region operate under the Vessel Day Scheme - which takes into account shifts in stocks and catches making it a good example where a management system has been set up that is somewhat anticipatory [@Aqorau:2018bh]. However, a low vulnerability index does not necessarily translate to no-risk of international conflict or threat to stocks sustainability. While it gives countries more time to prepare for substantial changes, stocks with relatively small overall participation on catches or revenue can destabilize cooperative shared stock management agreement as shown by the Atlantic mackerel example [@Spijkers:2017ij, Chapter 1].

@@

Changes in SSR will be seen across the world, suggesting that climate change impacts on transboundary stock management is a global problem (Figure \@ref(fig:toeFigure5)). Poleward shifts in SSR are expected in regions of the world with relatively large EEZs and few boundaries, like the North East Pacific. However, despite observations and modelling work indicating that, generally, climate change is shifting stocks distributions poleward [@Cheung:2010dt; @Poloczanska:2016kk; @IPCC:2019tn; @Fogarty:2017eo], the Eastern Tropical Pacific will see EEZs from Mexico to Ecuador losing SSR to their southern neighbors. Such range shifts towards the equator reflect movements to cooler habitats characteristic of equatorial upwelling systems. They are consistent with regional modeling studies suggesting that by mid $21^{st}$ century, marine stocks will move between 50 and 100 km southward of Mexico’s EEZ towards Peru under a high emission climate change scenario [@Clarke:2020ep].  Moreover, regions characterized by multiple political boundaries, like the Caribbean and the North East Atlantic, will see a complex network of changes in SSR with no clear pattern (Figure \@ref(fig:toeFigure5)B). These regions are known for being highly connected by both fish adult migration [@Levin:2018in] and larvae dispersal [@Ramesh:2019va] supporting the implementation of joint management plans [@Baudron:2020dj] and the call for new implemented shared policies [@Levin:2018in]. However, game theory suggests that the greater the number of negotiating parties , the harder it is for parties to reach an agreement [@Sumaila:2013hv], thus, making it particularly challenging for countries in these regions to coordinate the management of shifting shared stocks.

@@

<!-- ## Supplements -->

```{r toeFigureS1, eval = F,  echo = F, fig.width=9, fig.height=4, fig.cap =  "World Exclusive Economic Zones used in this study as devided by the Sea Around Us and their centroids (points)."}

@@

```{r toeFigureS2, eval = F,  echo = F, fig.width=9, fig.height=4, fig.cap = "Graphical representation of the transboundary index (TI) to determine the time of emergence of transboundary stocks. The index is based on the distance between the distributional centroid of the transboundary stock and the geographic centroid of the neighbouring Exclusive Economic Zones sharing the stock (Top panel). Time of emergence is defined as the first year when the TI overpasses historical values (Bottom panel)."}

@@

```{r toeFigureS3, eval = F,  echo = F, fig.width=4, fig.height=4, fig.cap =  "Average Time  of Emergence and standar dviation for a 64% confidence treshold  (purple) and a 98% confidence teshold (yellow)"}
